---
output:
  html_document:
    fig_caption: yes
    fig_height: 8
    fig_width: 12
    highlight: tango
    keep_md: yes
    number_sections: yes
    theme: cerulean
    toc: yes
  pdf_document:
    toc: yes
---

----------------------

###Coursera - Reproducible Research  
Duration 4 weeks / July-Aug2015  

[Coursera Assignment2 - Storms][1]  
Dataset is records for storm-events, for USA, from year 1950-2011nov.  
[NOAA Storm Database:] [2] verylarge CSV bzip2 [47Mb]  
[Instructions / Variable Name][3] for PDF file-download  
[FAQ][4] for PDF file download  

[1]: https://class.coursera.org/repdata-031/human_grading/view/courses/975144/assessments/4/submissions  
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[4]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

----------------------

###Instructions for Assignment

Document Layout  
Title: Your document should have a title that briefly summarizes your data analysis 
Synopsis: describes and summarizes your analysis in at most 10 complete sentences.  

section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis.
In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.  

section titled Results in which your results are presented.  
At least one figure containing a plot/ no more than three figures.  
You must show all your code for the work in your analysis document.echo = TRUE  
_

Submission of Output/ Code / Analysis Results:
Final rendered HTML output at Rpubs:
Reproducible study available at Github:

----------------------

#Title: Impact of weather-events, Harmful for Humans and the Economic consequences
####Analysis of NOAA Storm Database record (1950-2011). Region: USA.

======================

##Synopsis:

1. The (USA) NOAA Storm Database is a comprehensive record of Storm and related weather phenomena, along with additional data-points that support analysis of the consequenses of the weather-events on humans and the economic impact.

2. The records are available for duration 1950-2011; Though for many initial years only partial / incomplete information is there, no data has been disregarded and records are considered for calculation in-toto and as-available.

3. Much of the records are manual-transcribes, dataset is not tidy and the Variable Definitions not scientific; Hence requiring much pre-processing and cleaning of the dataset.

4. The money-value information is not available directly, through processing it is derived from two columns. Simple calculations to gets $ value as-is, not adjusted for time-value.

5. All dates are converted to POSITxct format with UCT as the common timezone.

6. EVTYPE is the event type; a best-guess has been used to understand the field_names and has been Categorised to 13 level, these will be fundamental to our further analysis.

7. Exploratory analysis of the dataset was done, for drawing-up a stratergy to process information needed and graphically plot the results - which visually answer the assignment-questions.

8. This report submits an Analysis based on information from the Dataset. It does not critically analyse the available records, extrapolate trends, nor outlines suggestions for response, and yet provides an office adequate information to decide on allocation and prioritization of resources.

----------------------

##Pre-setting

```{r 1}
setwd("D:/VIVEK/DataAnalysis/Coursera/5 RepRes/RepRes_Assignment2_Storms"); getwd()

require(stringr,  quietly = TRUE)
require(lubridate,  quietly = TRUE)
require(reshape2,  quietly = TRUE)
require(grid,  quietly = TRUE)
require(gridExtra,  quietly = TRUE)
require(ggplot2,  quietly = TRUE)
require(scales, quietly = TRUE)
require(knitr,  quietly = TRUE)
require(dplyr,  quietly = TRUE)

sessionInfo()
#Version 0.99.442 - © 2009-2015 RStudio, Inc.
```

## Knitr Global Options
```{r 2}
opts_chunk$set(echo=TRUE, eval=TRUE, results='as.is', cache= FALSE, strip.white= TRUE, tidy = TRUE)
```

----------------------

##Data Processing

###Loading the data

Sets the paths to the dataset/ files
```{r 3}
webURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipfilepath <- "repdata_data_StormData.csv.bz2"
rdsfilepath <- "StormData.RDS"
```
-

Checks if the file is avalable in local folder / If not downloads the zipfile
```{r 4}

if (!file.exists(zipfilepath)) {
	message("wait for file to be downloaded from website-url")
	download.file(url = webURL, destfile = zipfilepath) # not used method = "curl"
}
```
-

Checks if the data has been uploaded/ If not extracts data from the zipfile directly 
If data was already uploaded/ look for RDS file [As filesize is large, cache is necessary]
If available, reads file from RDS (much faster, and is the option used to knit this report)

```{r 5}
RDSloaded <- FALSE
message("main data set is being loaded")
if (!file.exists("rdsfilepath")) {
	message("large file...do patiently wait...will take many minutes")
	mdata <- read.csv(file = bzfile(zipfilepath), strip.white = TRUE)
	saveRDS(mdata, file = "rdsfilepath")
} else {
	message("extracting from compressed file and reading dataset into environment")
	mdata <- readRDS("rdsfilepath")
	RDSloaded <- TRUE
}
```

----------------------

###Explore the Dataset

Knowing the dataset and understanding the variable provides scope
Build a stratergy sequence to process the dataset

```{r 6}
dim(mdata) #ncol= records #nrow = variables
str(mdata); names(mdata) #head(mdata, n=3)
#issues: large dataset, not tidy, manual record entry, arbitarery naming.
```
-

Pictoral Exploration: provides a quick visual of the large datatable.
Graphical plots invisible for report/ used for understanding only.

```{r 7, eval= FALSE, results ='hide'}
plot(table(mdata$STATE__)) #
plot(log(table(mdata$TIME_ZONE))) #
plot(table(mdata$STATE)) #
plot(log(table(mdata$EVTYPE))) #
plot(log(table(mdata$FATALITIES))) #
plot(log(table(mdata$INJURIES))) #
table(mdata$PROPDMGEXP) #
table(mdata$CROPDMGEXP) #
```

----------------------

###Cleaning the data
The datset is not tidy / List_levels not defined
Lot of manual entry / spelling mistakes
Many fields of of Blank data
Wrong / irrelevant / incomprehensible entries

###Data Processing 1
The USD value is in two seperate columns; one with the numerical value; second the multiplier
```{r 8}
message("these code-lines take a few minutes to calculate")
multifunction <- function(unit, multiplier) unit * switch(toupper(multiplier), H=100, K=1000, M=1000000, B=1000000000, 1)
mdata$propDamage <- with(mdata, mapply(multifunction, PROPDMG, PROPDMGEXP))
mdata$cropDamage <- with(mdata, mapply(multifunction, CROPDMG, CROPDMGEXP))
message("two columns added to maindata for PropertyDamage and CropDamage in numerical_USD_Amount")
```
-

###Data Processing 2
convert dates to POSIXct format
```{r 9}
str(mdata$BGN_DATE)
mdata$dateUTC <- mdy(str_extract(mdata$BGN_DATE, "[^ ]+"))
message("column added for Date as POSIXct")
```
-

###Data Processing 3
cleaning the EVTYPE and making Category
EVTYPE column retained; new EVTYPEcat column added
```{r 10}
str(mdata$EVTYPE) #levels(mdata$EVTYPE)
levels(mdata$EVTYPE)<- tolower(levels(mdata$EVTYPE))
mdata$EVTYPEcat <- mdata$EVTYPE
```
-

###Data Processing 4
Important: EVTYPE categories are defined

1. "lightning"		2."rain_shower" 		3."tornado_landwinds" 		4."typhoon_seawinds" 

5."fire & smoke"	6."low_visibility"		7."tsunami_oceansurge"		8."heat_temparature"

9."volcanic_activity"	10."flood, erosion & avalanche"	11."unknown condition"	 12."others"

note: there is no documentation available from the dataset authorities for this categorisation step Data Processing
```{r 11}
levels(mdata$EVTYPEcat)[grepl("lightning |lighting|lightning|ligntning| lightning| ligntning|lightning.|lighting",levels(mdata$EVTYPEcat),ignore.case = T)] <- "lightning"
levels(mdata$EVTYPEcat)[grepl("thunder|rain|hail|wet|downburst|precip|precipatation|shower|microburst", levels(mdata$EVTYPEcat),ignore.case = T)] <- "rain_shower"
levels(mdata$EVTYPEcat)[grep("snow|winter|wintry|blizzard|glaze|hail|spout|sleet|cold|ice|freez|icy|frost",levels(mdata$EVTYPEcat),ignore.case = T)] <- "winter_conditions"
levels(mdata$EVTYPEcat)[grepl("tornado|torndao|wnd|wind|gustnado|funnel",levels(mdata$EVTYPEcat),ignore.case = T)] <- "tornado_landwinds"
levels(mdata$EVTYPEcat)[grepl("typhoon|swells|storm|hurricane|tropical +storm|turbulence",levels(mdata$EVTYPEcat),ignore.case = T)] <- "typhoon_seawinds"
levels(mdata$EVTYPEcat)[grepl("fire|smoke",levels(mdata$EVTYPEcat),ignore.case = T)] <- "fire & smoke"
levels(mdata$EVTYPEcat)[grepl("fog|visibility|dark|dust",levels(mdata$EVTYPEcat),ignore.case = T)] <- "low_visibility"
levels(mdata$EVTYPEcat)[grepl("marine| surf|surge|tide|tstm|tsunami|current|rough + seas|wave|depression| rapidly rising water| seas",levels(mdata$EVTYPEcat),ignore.case = T)] <- "tsunami_oceansurge"
levels(mdata$EVTYPEcat)[grepl("heat|high +temp|temperature|record +temp|warm|dry|hot",levels(mdata$EVTYPEcat),ignore.case = T)] <- "heat_temparature"
levels(mdata$EVTYPEcat)[grepl("volcan",levels(mdata$EVTYPEcat),ignore.case = T)] <- "volcanic_activity"
levels(mdata$EVTYPEcat)[grepl("avalance|avalanche|flooding|fld|stream|+flood|slide|mud|dam|flash|landslump|erosion|erosin|rapidly rising water",levels(mdata$EVTYPEcat),ignore.case = T)] <- "flood, erosion & avalanche"
levels(mdata$EVTYPEcat)[grepl("summary|southeast|vog|none|northern|\\?|other|urban|small|criteria|apache|floyd",levels(mdata$EVTYPEcat),ignore.case = T)] <- "unknown condition" 
levels(mdata$EVTYPEcat)[grepl("wallcloud|county|record+low|excessive|high|seiche|heavy mix|excessive|no severe weather",levels(mdata$EVTYPEcat),ignore.case = T)] <- "unknown condition"
levels(mdata$EVTYPEcat)[grepl("hyperthermia/exposure|drowning|unseasonal low temp|driest|record low|unseasonably cool|cool spell|drought|large wall cloud|record cool|mild pattern|wall cloud",levels(mdata$EVTYPEcat),ignore.case = T)] <- "others"
```
-

Important: The inital dataset had 985 factors for EVTYPE; it is now reduced to 13 categories as stored in EVTYPEcat

```{r 12}
str(mdata$EVTYPEcat)
levels(mdata$EVTYPEcat) 
table(mdata$EVTYPEcat)
```
-

###Data Processing 4
Make dataset lean, improve speed of system:
Purge many columns from the dataset, which will not be used in analysis.
```{r 13}
mdata<-select(mdata, STATE__, STATE, EVTYPE, FATALITIES, INJURIES, propDamage, cropDamage, dateUTC, EVTYPEcat)
names(mdata)
```

----------------------

## Exploratory Analysis
Graphical plots invisible for report/ used for understanding only.
```{r 14, echo=TRUE, eval = FALSE, results ='hide'}
invisible(hist(mdata$dateUTC, breaks = 61))
invisible(plot(mdata$dateUTC,mdata$FATALITIES))
invisible(plot(mdata$dateUTC,mdata$INJURIES))
invisible(pairs(EVTYPE ~ FATALITIES + INJURIES + propDamage + cropDamage,
	  data = mdata,  main = "pairs plot", 
	  subset = FATALITIES >0 & INJURIES >0 & propDamage >0 & propDamage>0))
invisible(plot(mdata$EVTYPE,mdata$propDamage))
invisible(plot(mdata$EVTYPE,mdata$cropDamage))
```
-

Understanding the Dataset; the number of events/ ratios
```{r 15}

t<-nrow(mdata)							#Ans: total recorded incidents [1] 902,297

dl<-sum(mdata$FATALITIES) 				#Ans: death_count [1] 15,145

icd<-sum(mdata$FATALITIES >0, na.rm=TRUE)#Ans: fatality_incidents [1] 6,974

jl<-sum(mdata$INJURIES)   				 #Ans: injury_count [1] 140,528

ici<-sum(mdata$INJURIES >0, na.rm=TRUE)	#Ans: injury_incidents [1] 17,604

pl<-sum(mdata$propDamage) 				 #Ans: Total Dollar Value [1] 427,318,652,972

icp<-sum(mdata$propDamage >0, na.rm=TRUE)#Ans: propertyloss_incidents [1] 239,174

cl<-sum(mdata$cropDamage) 				 #Ans: Total Dollar Value [1] 49,104,192,181

icc<-sum(mdata$cropDamage >0, na.rm=TRUE)#Ans: croploss_incidents [1] 22,099
```
-

Ratio of number of Human_Incident to total recorderd Incidents
```{r 16}

t/icd #1 in every 130, recorded incidents has led to atleast 1 fatality

t/ici #1 in every 51, recorded incidents has led to atleast 1 injury
```
-

Ratio of number of Damage_Incident to total recorderd Incidents
```{r 17}
t/icp #1 in every 3, recorded incidents has caused property damage

t/icc #1 in every 40, recorded incidents has caused crop damage
```

Average per incident 
```{r 18}
#fatality_incident
dl/icd				#Ans: 2.17 average deaths per incident

#injury_incident
jl/ici				#Ans: 7.98 average persons injured per incident

#propertyloss_incident
pl/icp				#Ans: $1,786,643 average loss per incident

#croploss_incident
cl/icc				#Ans: $2,222,010 average loss per incident
```

----------------------

caution note: for below code block, group_by does not work if 'plyr'is loaded after 'dplyr
search() #display the packages detach("package:dplyr", unload=TRUE) #unload the package

----------------------

##Analysis:

###Which categories are most harmful?
Impact on Human Life (Fatalities and Injuries)

####Analysis Code 1
Analysing the effect of Event-type-categories on human_incidents (deaths and injuries)
```{r 19}
harmful <- filter(mdata, FATALITIES > 0 & INJURIES > 0)
harmful <- dplyr::group_by(harmful, EVTYPEcat)
harmful <- dplyr::summarise(harmful, FATALITIES=sum(FATALITIES, na.rm = T), INJURIES=sum(INJURIES, na.rm = T))
harmful <- dplyr::arrange(harmful, desc(FATALITIES+INJURIES))
names(harmful) <- c("EVTYPEcat","human_deaths","human_injuries")
meltharmful <- melt(harmful[1:4,], id.vars = "EVTYPEcat", measure.vars = c("human_deaths","human_injuries"))
```
-

####Presentation Code 1
The ggplot grapplots only the top four EVTYPEcat / as the rest comparitively have less impact
```{r 20}
plot1 <-ggplot(data = meltharmful, aes(x=reorder(EVTYPEcat, desc(value)), y=value/10^3, fill=variable)) +
	geom_bar(stat = "identity", width = 0.8, position = "dodge") +
	theme(axis.text.x = element_text(angle = 60, hjust = 1))+
	ylab("Number of Person  ( in Thousand)") +
	xlab("Major Event Category") +
	labs(fill = "Incidents") +
	theme(legend.position= c(0.8,0.8), legend.background = element_rect(fill="transparent"))+
	ggtitle("Top 4 Weather Events | Dangerous for Human Life") +
	theme(plot.title = element_text( size=12, lineheight=.6, face="bold"))+
	geom_text(aes(label = value), size =3, hjust = 1.2, vjust = -0.8, alpha = 0.4)
```
-
### Record the year of events, for the top categories - harmful incident.

####Analysis Code 2
Subset the primary cleandataset for the four category variables
```{r 21}
harmloc <- rbind(filter(mdata, EVTYPEcat == "tornado_landwinds") ,filter(mdata, EVTYPEcat == "heat_temparature"),filter(mdata, EVTYPEcat == "winter_conditions"),filter(mdata, EVTYPEcat == "flood, erosion & avalanche")) %>% droplevels()
```
-

####Presentation Code 2
The ggplot plots both deaths and injuries
```{r 22}
plot2 <-ggplot(harmloc, aes(x = dateUTC, y = INJURIES, group = EVTYPEcat)) +
	geom_line(aes(stat = "identity", colour="human_injuries"), alpha= 0.2)+
	geom_line(data = harmloc,aes(y = FATALITIES, stat = "identity", colour="human_deaths"), alpha= 0.5)+
	scale_y_continuous(limits = c(0,400), oob=rescale_none) +
	facet_wrap( ~ EVTYPEcat, nrow = 1) +
	coord_flip()+
	xlab("time-line.years 1950-2011") +
	ylab("quantity of casuality") +
	ggtitle("Timeline: Annual Casuality Incidents. EVTYPE by Panel - for the Top 4 Weather Events - dangerous for humanlife") +
	theme(plot.title = element_text( size=10, lineheight=.6, face="bold")) +
	theme(legend.position= c(0.9,0.3), legend.background = element_rect(fill="transparent"))+
	theme(legend.title = element_text(colour="darkgrey", size=10, face="bold"))+
	scale_color_discrete(name="Incidents")
``````
-

###Which categories are most destructive?
Economic Consequences (Property and Crop Damage)

####Analysis Code 3
```{r 23}
damage <- filter(mdata, propDamage > 0 & propDamage > 0)
damage <- dplyr::group_by(damage, EVTYPEcat)
damage <- dplyr::summarise(damage, propDamage=sum(propDamage, na.rm = T), cropDamage=sum(cropDamage, na.rm = T))
damage <- dplyr::arrange(damage, desc(propDamage+cropDamage))
meltdamage <- melt(damage[1:5,], id.vars = "EVTYPEcat", measure.vars = c("propDamage","cropDamage"))
```
-

####Presentation Code 3
The ggplot grapplots only the top five EVTYPEcat / as the rest comparitively have less impact
```{r 24}
plot3 <-ggplot(data = meltdamage, aes(x=reorder(EVTYPEcat, desc(value)), y=value/10^9, fill=variable)) +
	geom_bar(stat = "identity", width = 0.8, position = "dodge") +
	ylab("Financial Measure: Loss in $ (Billion)") +
	xlab("Major Event Category") +
	labs(fill = "Legend") +
	theme(legend.position= c(0.8,0.8), legend.background = element_rect(fill="transparent"))+
	ggtitle("Top 5 Weather Events | causing Economic Damage") +
	theme(plot.title = element_text( size=12, lineheight=.6, face="bold"))+
	geom_text(aes(label = round(value/10^9)), size =3, hjust = 1.1, vjust = -0.8, alpha = 0.4)+
	theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

----------------------

##Results

###Combined Plots in single Graphical Output

Setting up the multiplot function  
Code-Reference: [R-cookbook][5]
[5] http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_%28ggplot2%29/
```{r 25}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```
-

####Presentation Code 3

Graphical Output: Single figure with Three plots and multi-pane [Analysis of data in visible form]
This meets the assignemtnt instructions to have less than three plots in the report
```{r 26}
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 2)))
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)
print(plot2, vp = vplayout(1, 1:2))  # key is to define vplayout
print(plot1, vp = vplayout(2:3, 1))
print(plot3, vp = vplayout(2:3, 2))
```

----------------------

###Q1. and answer

Which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?[Across the United States]

Ans1:
From the output, we find the harmful for humans (fatalities + injuries) 'Top Five Categories' are 1. tornado_landwinds 2. heat_temparature 3. winter_conditions 4. flood, erosion & avalanche
The harm-impact of the other 9 categories is comparitively negligible, hence not plotted.
-

###Q2. and answer

Which types of events have the greatest economic consequences?[Across the United States]

Ans2:
From the output, we find the destructive (propDamage + cropDamage) 'Top Five Categories' are 1.tornado_landwinds 2.flood, 3.erosion & rain_shower  4.winter_conditions  5. fire&smoke
The harm-impact of the other 10 categories is comparitively negligible, hence not plotted.
-

Note: Analysing for the Categories, there is a key-obervation to be made.
Records for tornado_landwinds available since 1950 | for the others there is minimal records and mostly not available till 1996

End of Report.
----------------------